Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Community Detection and Parameter Estimation

### 2021-06-30 summary.tex
Update paper

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{indep}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}
\newcommand{\BG}{\text{BernoulliGraph}}
\newcommand{\PABM}{\text{PABM}}
\newcommand{\RDPG}{\text{RDPG}}
\newcommand{\GRDPG}{\text{GRDPG}}
\newcommand{\Multinomial}{\text{Multinomial}}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{definition}
\newtheorem*{example}{Example}
\newpage
\spacingset{1.5} % DON'T change the spacing!

@@

Statistical analysis on graphs or networks often involves the
partitioning of a graph into disconnected subgraphs or clusters. This is
often motivated by the assumption that there exist underlying and
unobserved communities to which each vertex of the graph belongs, and
edges between pairs of vertices are determined by drawing from a
probability distribution based on the community relationships between
each pair. The goal of this analysis then is population community
detection, or the recovery of the true underlying community labels for
each vertex, up to permutation, with some additional parameter
estimation being of possible interest, assuming some underlying
probability model. One such model is the Stochastic Block Model (SBM),
first proposed by \citet{doi:10.1080/0022250X.1971.9989788}, which
assumes that the edge probability from one vertex to another follows a
Bernoulli distribution with fixed probabilities for each pair of
community labels. Other random graph models have been proposed and
studied, such as the Degree-Corrected Block Model (DCBM), introduced by
\citet{Karrer_2011}, which is a generalization of the SBM. The
Popularity Adjusted Block Model (PABM) was then introduced by
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the
DCBM to address the heterogeneity of edge probabilities within and
between communities while still maintaining distinct community
structure.

@@

The Random Dot Product Graph (RDPG) model proposed by
\citet*{10.1007/978-3-540-77004-6_11} is another graph model with
Bernoulli edge probabilities. Under this model, each vertex of the graph
can be represented by a point in some latent space such that the edge
probability between any pair of vertices is given by their corresponding
dot product in the latent space, i.e., given a latent positions
\(x_1, ..., x_n \in \mathbb{R}^d\), the edge probability matrix is
\(P = X X^\top\) where
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\). The assortative
SBM is equivalent to a special case of the RDPG for which all vertices
of a given community share the same position in the latent space
\cite{lyzinski2014}. It has also been shown that similar random graph
models, including the DCBM, can be represented in a similar fashion
\cite{lyzinski2014, rubindelanchy2017consistency}. An analogous
property exists for the PABM, not for the RDPG model but the
\emph{Generalized} Random Dot Product Graph (GRDPG) model, which is an
extension of the RDPG to allow edge probability matrices that are not positive
semidefinite. This relationship will be explored in this paper and exploited to
construct algorithms for community detection and parameter estimation for the
PABM.

\hypertarget{notation}{%
\subsection{Notation and Scope}\label{notation}}

Let $G = (V, E)$ be an unweighted, undirected, and hollow graph with vertex set $V$ ($|V| = n$) and edge set $E$. $A \in \{0, 1\}^{n \times n}$ represents the adjacency matrix of $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $0$ otherwise. Because $G$ is symmetric and hollow,
$A_{ij} = A_{ji}$ and $A_{ii} = 0$ for each $i, j \in [n]$. We further restrict our analyses to Bernoulli graphs. Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $1 \leq i < j \leq n$  (setting $A_{ji} = A_{ij}$ and $A_{ii} = 0$). We denote $A \sim \BG(P)$ as graph with adjacency matrix $A$ sampled from edge probability matrix $P$ in this manner. If each vertex has a hidden label in $[K]$, they are denoted as $z_1, ..., z_n$. Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the collection of $n$ latent vectors $x_1, ..., x_n \in \mathbb{R}^d$.

@@

In this section, we show that the PABM is a special case of the GRDPG, i.e.,
graph $G$ drawn from the PABM can be represented by a collection of latent
vectors in Euclidean space. We further show that the latent configuration that
induces the PABM consists of orthogonal subspaces with each subspace
corresponding to a community.

\hypertarget{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}{%
\subsection{The Popularity Adjusted Block Model and the Generalized
Random Dot Product
Graph}\label{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}}

\begin{definition}[Popularity Adjusted Block Model]
\label{pabm}
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a
graph $G = (V, E)$ with adjacency matrix $A$ such that $A \sim \BG(P)$.
Let each vertex have a community label between $1$ and $K$.
Then $G$ is drawn from a Popularity Adjusted Block Model if
each vertex has $K$ popularity parameters that describe its affinity
toward each of the $K$ communities, i.e., vertex $i$ has popularity parameters
$\lambda_{i1}, ..., \lambda_{iK}$, and
each $P_ij = \lambda_{i z_j} \lambda_{j z_i}$.

Another characterization of the PABM is as follows.
Let the rows and columns of $P$ be arranged by community label
such that $n_k \times n_l$ block $P^{(kl)}$
describes the edge probabilities between vertices in communities
$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$).
If each block $P^{(kl)}$ can be written as the outer product of two vectors:

@@

for a set of $K^2$ popularity vectors $\{\lambda^{(st)}\}_{s, t = 1}^K$ where each
$\lambda^{(st)}$ is a column vector of dimension $n_s$, then graph $G$ is drawn from a PABM with parameters $\{\lambda^{(st)}\}_K$ if its its corresponding adjacency matrix $A \sim \BG(P)$.
\end{definition}

We will use the notation \(A \sim \PABM(\{\lambda^{(kl)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\lambda^{(kl)}\) consisting of \(K\) underlying communities.

\begin{definition}[Generalized Random Dot Product Graph]
\label{grdpg}
Let graph $G = (V, E)$ be drawn as $A \sim \BG(P)$.
If $\exists X \in \mathbb{R}^{n \times d}$ such that

\begin{equation} \label{eq:grdpg}
  P = X I_{p,q} X^\top
\end{equation}

for some $d, p, q \in \mathbb{N}$ and $p + q = d$, then
$G$ is drawn from the Generalized Random Dot Product Graph with latent positions $x_1, ..., x_n \in \mathbb{R}^d$ and signature $(p, q)$.
\end{definition}

We will use the notation \(A \sim \GRDPG_{p,q}(X)\) to denote a random
adjacency matrix \(A\) drawn from latent positions \(X\) and signature
\((p, q)\). If instead of fixed latent positions, they are drawn $X_1, ..., X_n \iid F$, we denote the GRDPG as $(A, X) \sim \GRDPG_{p,q}(F, n)$.

\begin{definition}[Indefinite Orthogonal Group]
The indefinite orthogonal group with signature $(p, q)$ is
the set $\{Q \in \mathbb{R}^{d \times d} : Q I_{p, q} Q^{\top} = I_{p, q}\}$,
denoted as $\mathbb{O}(p, q)$.
\end{definition}

\begin{remark}
Like the RDPG, the latent positions of a GRDPG are not unique
\cite{rubindelanchy2017statistical}.
More specifically, if $P_{ij} = x_i^\top I_{p, q} x_j$, then we also have for any
$Q \in \mathbb{O}(p, q)$,
$(Q x_i)^\top I_{p, q} (Q x_j) = x_i^\top (Q^\top I_{p, q} Q) x_j =
x_i^\top I_{p, q} x_j = P_{ij}$.
Unlike in the RDPG case, transforming the latent positions via multiplication
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or
distances.
\end{remark}

@@

We previously stated one possible set of latent positions that result in
the edge probability matrix of a PABM, \(P = (XU) I_{p, q} (XU)^\top\). If
we have (or can estimate) \(XU\) directly, then both the community
detection and parameter identification problem are trivial since \(U\)
is orthonormal and fixed for each value of \(K\). However, direct
identification or estimation of \(XU\) is not possible
\cite{rubindelanchy2017statistical}.

If we decompose \(P = Z I_{p, q} Z^\top\), then
\(\exists Q \in \mathbb{O}(p, q)\) such that \(XU = Z Q\). Even if we
start with the exact edge probability matrix, we cannot recover the
``original'' latent positions \(XU\). Note that unlike in the case of
the RDPG, \(Q\) is not necessarily an orthogonal matrix. If \(z_i\)'s
are the rows of \(XU\), then
\(||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2\), and
\(\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle\). This
prevents us from using the properties of \(XU\) directly. In particular,
if \(Q \in \mathbb{O}(n)\), then we could use the fact that
\(\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0\) if
vertices \(i\) and \(j\) are in different communities.

@@

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\). Therefore, given
\(P^{(kl)}\), \(\lambda^{(kl)}\) and \(\lambda^{(lk)}\) are solvable
up to multiplicative constant using singular value
decomposition. More specifically, 
let \(P^{(kl)} = (\sigma^{(kl)})^2 u^{(kl)} (v^{(kl)})^\top\)
be the singular value decomposition of \(P^{(kl)}\).
\(u^{(kl)} \in \mathbb{R}^{n_k}\) and 
\(v^{(kl)} \in \mathbb{R}^{n_l}\) are vectors
and \(\sigma^{(kl)}\) is a scalar. Then \(\lambda^{(kl)} = s_1 u^{(kl)}\)
and \(\lambda^{(lk)} = s_2 v^{(kl)}\) 
for unidentifiable $s_1 s_2 = (\sigma^{(kl)})^2$.
Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators by taking the SVD of each $A^{(kl)}$. 
Since each $\lambda^{(kl)}$ is not strictly identifiable,
we instead estimate each $\tilde{\lambda}^{(kl)} = \sigma^{(kl)} u^{(kl)}$.

@@

For each simulation, community labels are drawn from a multinomial
distribution, the popularity vectors \(\{\lambda^{(kl)}\}_K\) are drawn
from two types of joint distributions depending on whether \(k = l\),
the edge probability matrix \(P\) is constructed using the popularity
vectors, and finally an unweighted and undirected adjacency matrix \(A\)
is drawn from \(P\). OSC is then used for community detection, and this
method is compared against SSC \cite{noroozi2019estimation,soltanolkotabi2014}
and MM \cite{igraph}
\cite{307cbeb9b1be48299388437423d94bf1}. True community labels are used
with Algorithm 3 to estimate the popularity vectors
\(\{\lambda^{(kl)}\}_K\), and this method is then compared against an
MLE-based estimator described by \citeauthor{noroozi2019estimation} and
\citeauthor{307cbeb9b1be48299388437423d94bf1}.

Modularity Maximization is NP-hard, so
\citeauthor{307cbeb9b1be48299388437423d94bf1} used the Extreme Points
(EP) algorithm \cite{le2016}, which is \(O(n^{K - 1})\), as a greedy
relaxation of the optimization problem. For these simulations, the
Louvain algorithm was used for modularity maximization,
as \citeauthor{307cbeb9b1be48299388437423d94bf1}'s implementation proved to be
prohibitively computationally expensive for \(K > 2\). For \(K = 2\), it
was verified that the Louvain algorithm produces comparable results
to EP-MM.

@@

We also define two types of parameter estimation error. 
First, we estimate the popularity vectors directly and compute the RMSE. 
In this case, the "true" popularity vectors are derived from 
taking the SVD of each edge probability block $P^{(kl)}$ 
to avoid the unidentifiable multiplicative constants.

\[\text{RMSE}(\{\hat{\lambda}^{(kl)}\}_K, \{\lambda^{(kl)}\}_K) = 
\sqrt{\frac{1}{n} \sum_{k < l} \min_{s = \pm 1} 
\|s \hat{\lambda}^{(kl)} - \lambda^{(kl)} \|_2^2}\]

We can also avoid the unidentifiable multiplicative constant more directly 
by reconstructing each 
$\hat{P}^{(kl)} = \hat{\lambda}^{(kl)} (\hat{\lambda}^{(lk)})^\top$, 
which we use to define another parameter estimation error.

\[\text{RMSE}(\hat{P}, P) =
\sum_{k,l} \sqrt{\frac{1}{n_k n_l} \|P^{(kl)} - \hat{P}^{(kl)}\|_F^2}\]

@@

In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
multinomial distribution with mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(kl)}\}_K\) according
to the drawn community labels, \(P\) was constructed using the drawn
\(\{\lambda^{(kl)}\}_K\), and \(A\) was drawn from \(P\) by
\(A_{ij} \indep \Bernoulli(P_{ij})\). Each simulation has
a unique edge probability matrix \(P\).

@@

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\), (i.e.,
  each community label has an equal probability of being drawn)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim} \Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim} \Betadist(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(kl)} \stackrel{\text{iid}}{\sim} \Betadist(1, 2)\) for
  \(k \neq l\)
\end{itemize}

@@

Given ground truth community labels, Algorithm 3 and the MLE-based
plug-in estimators \cite{307cbeb9b1be48299388437423d94bf1} 
perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\).

@@

\begin{lemma}
\label{lemma1}
Let $P = V D V^\top$ be the spectral decomposition of the edge probability
matrix for a PABM. Then $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is
defined as in (\ref{eq:xy}).
\end{lemma}

\emph{Proof}. By Theorem 2, \(P = X U I_{p, q} U^\top X^\top\), where
\(X\) is defined as in (\ref{eq:xy}) and $p = K(K+1)/2$ and $q = K(K-1)/2$. % are given by $p =
% K(K+1)/2$ and $q = are defined as
% in equations (\ref{eq:p}) and (\ref{eq:q}).
Alternatively, the spectral
decomposition can be written as
\(P = V D V^\top = V |D|^{1/2} I_{p, q} |D|^{1/2} V^\top\) for the same
\((p, q)\) and \(|\cdot|^{1/2}\) is applied entry-wise. Thus for some
\(Q \in \mathbb{O}(p, q)\),

\[X U Q = V |D|^{1/2}\]

Therefore, using the fact that \(U U^\top = I\) and \(V^\top V = I\),

\[(V |D|^{1/2}) ((V |D|^{1/2})^\top (V |D|^{1/2}))^{-1} (V |D|^{1/2})^\top =
(X U Q) ((X U Q)^\top (X U Q))^{-1} (X U Q)^\top\]

The right-hand side becomes

\[\begin{split}
  (X U Q) ((X U Q)^\top (X U Q))^{-1} (X U Q)^\top &
    = X U Q Q^{-1} U^\top (X^\top X)^{-1}
    U (Q^\top)^{-1} Q^\top U^\top X^\top \\
  & = X U U^\top (X^\top X)^{-1} U U^\top X^\top \\
  & = X (X^\top X)^{-1} X^\top
\end{split}\]

The left-hand side becomes:

\[\begin{split}
  (V |D|^{1/2}) ((V |D|^{1/2})^\top (V |D|^{1/2}))^{-1} (V |D|^{1/2})^\top & =
    V |D|^{1/2} |D|^{-1/2} (V^\top V)^{-1} |D|^{-1/2} |D|^{1/2} V^\top \\
  & = V V^\top
\end{split}\]

\textbf{Proof of Theorem \ref{theorem3}}. By Lemma \ref{lemma1},
\(V V^\top = X (X^\top X)^{-1} X^\top\) where \(X\) is defined as in
(\ref{eq:xy}). Since \(X\) is block diagonal with each block
corresponding to one community, \(X (X^\top X)^{-1} X^\top\) is also a
block diagonal matrix with each block corresponding to a community and
zeros elsewhere. Therefore, if vertices \(i\) and \(j\) belong to
different communities, then the \(ij\)\textsuperscript{th} element of
\(n X (X^\top X)^{-1} X^\top = n V V^\top = B\) is 0.

>>>>

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\newcommand{\diag}{\text{diag}}
\newcommand{\tr}{\text{Tr}}
\newcommand{\blockdiag}{\text{blockdiag}}
\newcommand{\indep}{\stackrel{\text{indep}}{\sim}}
\newcommand{\iid}{\stackrel{\text{iid}}{\sim}}
\newcommand{\Bernoulli}{\text{Bernoulli}}
\newcommand{\Betadist}{\text{Beta}}
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\theoremstyle{definition}
\newtheorem*{example}{Example}
\newpage
\spacingset{1.5} % DON'T change the spacing!

@@

Statistical analysis on graphs or networks often involves the
partitioning of a graph into disconnected subgraphs or clusters. This is
often motivated by the assumption that there exist underlying and
unobserved communities to which each vertex of the graph belongs, and
edges between pairs of vertices are determined by drawing from a
probability distribution based on the community relationships between
each pair. The goal of the analysis then is population community
detection, or the recovery of the true underlying community labels for
each vertex, up to permutation (with some additional parameter
estimation being of possible interest), assuming some underlying
probability model. One such model is the Stochastic Block Model (SBM),
first proposed by \citet{doi:10.1080/0022250X.1971.9989788}, which
assumes that the edge probability from one vertex to another follows a
Beronulli distribution with fixed probabilities for each pair of
community labels. Other random graph models have been proposed and
studied, such as the Degree-Corrected Block Model (DCBM), introduced by
\citet{Karrer_2011}, which is a generalization of the SBM. The
Popularity Adjusted Block Model (PABM) was then introduced by
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the
DCBM to address the heterogeneity of edge probabilities within and
between communities while still maintaining distinct community
structure.

@@

The Random Dot Product Graph (RDPG) model proposed by
\citet*{10.1007/978-3-540-77004-6_11} is another graph model with
Bernoulli edge probabilities. Under this model, each vertex of the graph
can be represented by a point in some latent space such that the edge
probability between any pair of vertices is given by their corresponding
dot product in the latent space, i.e., given a latent positions
\(x_1, ..., x_n \in \mathbb{R}^d\), the edge probability matrix is
\(P = X X^\top\) where
\(X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top\). The SBM
is equivalent to a special case of the RDPG model in which all vertices
of a given community share the same position in the latent space
\cite{lyzinski2014}. It has also been shown that similar random graph
models, including the DCBM, can be represented in this way
\cite{lyzinski2014} \cite{rubindelanchy2017consistency}. An analogous
property exists for the PABM but not for the RDPG model but under the
\emph{Generalized} Random Dot Product Graph (GRDPG) model. This
relationship will be explored in this paper and exploited to construct
algorithms for community detection and parameter estimation for the
PABM.

In this paper, we will only consider undirected graphs, that is the edge
weight from vertex \(i\) to vertex \(j\) is equal to the edge weight in
the opposite direction, from vertex \(j\) to vertex \(i\). Furthermore,
we will only consider unweighted graphs with binary (\(0\) or \(1\))
edge weights We will also assume that graphs are hollow, i.e., there are
no edges from a vertex to itself. All such graphs can be represented by
a symmetric adjacency matrix \(A \in \{0, 1\}^{n \times n}\) for which
\(A_{ij} = 1\) if there exists an edge between vertices \(i\) and \(j\)
and \(0\) otherwise, and \(A\) is an element-wise independent Bernoulli
draw from a symmetric edge probability matrix
\(P \in [0, 1]^{n \times n}\).

@@

\hypertarget{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}{%
\subsection{The popularity adjusted block model and the generalized
random dot product
graph}\label{the-popularity-adjusted-block-model-and-the-generalized-random-dot-product-graph}}

\begin{definition}[Popularity Adjusted Block Model]
\label{pabm}
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl)}$ describes the edge probabilities between vertices in communities 
$k$ and $l$ ($P^{(lk)} = (P^{(kl)})^\top$). 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $\text{Bernoulli}(P)$, i.e., 
$A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

If each block $P^{(kl)}$ can be written as the outer product of two vectors:

@@

for a set of $K^2$ fixed vectors $\{\lambda^{(st)}\}_{s, t = 1}^K$ where each 
$\lambda^{(st)}$ is a column vector 
of dimension $n_s$, then graph $G$ and its corresponding adjacency matrix $A$ 
is a realization of a popularity adjusted block model with parameters 
$\{\lambda^{(st)}\}_{s, t = 1}^K$. 
\end{definition}

We will use the notation \(A \sim PABM(\{\lambda^{(kl)}\}_K)\) to denote
a random adjacency matrix \(A\) drawn from a PABM with parameters
\(\lambda^{(kl)}\) consisting of \(K\) underlying communities.

\begin{definition}[Generalized Random Dot Product Graph] 
\label{grdpg} 
Let $P \in [0, 1]^{n \times n}$ be a symmetric edge probability matrix for a 
set of $n$ vertices, $V$. If $\exists X \in \mathbb{R}^{n \times d}$ such that

\begin{equation} \label{eq:grdpg}
  P = X I_{pq} X^\top
\end{equation}

for some $d, p, q \in \mathbb{N}$ and $p + q = d$, then 
graph $G = (V, E)$ with adjacency matrix $A$ such that 
$A_{ij} \stackrel{\text{indep}}{\sim}\text{Bernoulli}(P_{ij})$ for $i > j$ ($A_{ij} = A_{ji}$ and 
$A_{ii} = 0$) is a draw from the generalized random dot product graph model 
with latent positions $X$ and signature $(p, q)$. More precisely, if vertices 
$i$ and $j$ have latent positions $x_i$ and $x_j$ respectively, then the edge 
probability between the two is $P_{ij} = x_i^\top I_{pq} x_j$, and $X$ contains 
the latent positions as rows $x_i^\top$. 
\end{definition}

We will use the notation \(A \sim GRDPG_{p,q}(X)\) to denote a random
adjacency matrix \(A\) drawn from latent positions \(X\) and signature
\((p, q)\).

\begin{definition}[Indefinite Orthogonal Group] 
The indefinite orthogonal group with signature $(p, q)$ is 
the set $\{Q \in \mathbb{R}^{d \times d} : Q I_{pq} Q^{\top} = I_{pq}\}$, 
denoted as $\mathbb{O}(p, q)$.
\end{definition}

\begin{remark}
Like the RDPG, the latent positions of a GRDPG are not unique 
\cite{rubindelanchy2017statistical}. 
More specifically, if $P_{ij} = x_i^\top I_{pq} x_j$, then we also have for any 
$Q \in \mathbb{O}(p, q)$, 
$(Q x_i)^\top I_{pq} (Q x_j) = x_i^\top (Q^\top I_{pq} Q) x_j = 
x_i^\top I_{pq} x_j = P_{ij}$. 
Unlike in the RDPG case, transforming the latent positions via multiplication 
by $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or 
distances. 
\end{remark}

@@

We previously stated one possible set of latent positions that result in
the edge probability matrix of a PABM, \(P = (XU) I_{pq} (XU)^\top\). If
we have (or can estimate) \(XU\) directly, then both the community
detection and parameter identification problem are trivial since \(U\)
is orthonormal and fixed for each value of \(K\). However, direct
identification or estimation of \(XU\) is not possible
\cite{rubindelanchy2017statistical}.

If we decompose \(P = Z I_{pq} Z^\top\), then
\(\exists Q \in \mathbb{O}(p, q)\) such that \(XU = Z Q\). Even if we
start with the exact edge probability matrix, we cannot recover the
``original'' latent positions \(XU\). Note that unlike in the case of
the RDPG, \(Q\) is not necessarily an orthogonal matrix. If \(z_i\)'s
are the rows of \(XU\), then
\(||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2\), and
\(\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle\). This
prevents us from using the properties of \(XU\) directly. In particular,
if \(Q \in \mathbb{O}(n)\), then we could use the fact that
\(\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0\) if
vertices \(i\) and \(j\) are in different communities.

@@

For any edge probability matrix \(P\) for the PABM such that the rows
and columns are organized by community, the \(kl\)\textsuperscript{th}
block is an outer product of two vectors, i.e.,
\(P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top\). Therefore, given
\(P^{(kl)}\), \(\lambda^{(kl)}\) and \(\lambda^{(lk)}\) are solvable
exactly (up to multiplication by \(-1\)) using singular value
decomposition. More specifically, let \(P^{(kl)} = \sigma^2 u v^\top\)
be the singular value decomposition of \(P^{(kl)}\).
\(u \in \mathbb{R}^{n_k}\) and \(v \in \mathbb{R}^{n_l}\) are vectors
and \(\sigma^2 > 0\) is a scalar. Then \(\lambda^{(kl)} = \pm \sigma u\)
and \(\lambda^{(lk)} = \pm \sigma v\). Given the adjacency matrix \(A\)
instead of edge probability matrix \(P\), we can simply use plug-in
estimators (algorithm 3), which converge to the true parameters.

@@

For each simulation, community labels are drawn from a multinomial
distribution, the popularity vectors \(\{\lambda^{(kl)}\}_K\) are drawn
from two types of joint distributions depending on whether \(k = l\),
the edge probability matrix \(P\) is constructed using the popularity
vectors, and finally an unweighted and undirected adjacency matrix \(A\)
is drawn from \(P\). OSC is then used for community detection, and this
method is compared against SSC \cite{noroozi2019estimation}
\cite{soltanolkotabi2014} and MM \cite{igraph}
\cite{307cbeb9b1be48299388437423d94bf1}. True community labels are used
with Algorithm 3 to estimate the popularity vectors
\(\{\lambda^{(kl)}\}_K\), and this method is then compared against an
MLE-based estimator described in \citeauthor{noroozi2019estimation} and
\citeauthor{307cbeb9b1be48299388437423d94bf1}.

Modularity Maximization is NP-hard, so
\citeauthor{307cbeb9b1be48299388437423d94bf1} used the Extreme Points
(EP) algorithm \cite{le2016}, which is \(O(n^{K - 1})\), as a greedy
relaxation of the optimization problem. For these simulations, the
Louvain algorithm was used, as the EP algorithm proved to be
prohibitively computationally expensive for \(K > 2\). For \(K = 2\), it
was verified that EP and Louvain produce comparable results.

@@

We also define parameter estimation error as the RMSE up to sign flip:

\[RMSE(\{\hat{\lambda}^{(kl)}\}, \{\lambda^{(kl)}\}) = 
\sqrt{\frac{1}{n} \sum_{k < l} \min_{s \in \{-1, 1\}} 
\sum_{i = 1}^{n_k} 
(s \hat{\lambda}^{(kl)}_i - \lambda^{(kl)}_i)^2}\]

@@

In each simulation, community labels \(z_1, ..., z_n\) were drawn from a
multinomial distribution with mixture parameters
\(\{\alpha_1, ..., \alpha_K\}\), then \(\{\lambda^{(kl)}\}_K\) according
to the drawn community labels, \(P\) was constructed using the drawn
\(\{\lambda^{(kl)}\}_K\), and \(A\) was drawn from \(P\) by
\(A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})\). Each simulation has
a unique edge probability matrix \(P\).

@@

\begin{itemize}
\tightlist
\item
  Number of vertices \(n = 128, 256, 512, 1024, 2048, 4096\)
\item
  Number of underlying communities \(K = 2, 3, 4\)
\item
  Mixture parameters \(\alpha_k = 1 / K\) for \(k = 1, ..., K\), (i.e.,
  each community label has an equal probability of being drawn)
\item
  Community labels
  \(z_k \stackrel{\text{iid}}{\sim}Multinomial(\alpha_1, ..., \alpha_K)\)
\item
  Within-group popularities
  \(\lambda^{(kk)} \stackrel{\text{iid}}{\sim}Beta(2, 1)\)
\item
  Between-group popularities
  \(\lambda^{(kl)} \stackrel{\text{iid}}{\sim}Beta(1, 2)\) for
  \(k \neq l\)
\end{itemize}

@@

Given ground truth community labels, Algorithm 3 and the MLE-based
plug-in estimators \cite{307cbeb9b1be48299388437423d94bf1}
\cite{noroozi2019estimation} perform similarly, with root mean square
error decaying at rate approximately \(n^{-1/2}\).

@@

\begin{lemma}
\label{lemma1}
Let $P = V D V^\top$ be the spectral decomposition of the edge probability 
matrix for a PABM. Then $V V^\top = X (X^\top X)^{-1} X^\top$ where $X$ is 
defined as in Eq.\eqref{eq:xy}.
\end{lemma}

\emph{Proof}. By Theorem 2, \(P = X U I_{p, q} U^\top X^\top\), where
\(X\) is defined as in Eq.~\eqref{eq:xy}), $p = K(K+1)/2$, and $q = K(K-1)/2$. % are given by $p =
% K(K+1)/2$ and $q = are defined as
% in equations (\ref{eq:p}) and (\ref{eq:q}). 
Alternatively we have the eigendecomposition \(P = V D V^\top = V
|D|^{1/2} I_{p, q} |D|^{1/2} V^\top\) and \(|\cdot|^{1/2}\) is applied entry-wise. Thus for some
\(Q \in \mathbb{O}(p, q)\),
\[X U Q = V |D|^{1/2}.\]
Let $R = U Q |D|^{-1/2}$. Then $R$ is {\em invertible} and hence
$$X (X^{\top} X)^{-1} X^{\top} = X R (R^{\top} X^{\top} X R)^{-1}
R^{\top} X^{\top}.$$ 
Furthermore $V = XR$ and since $V$ has orthonormal columns, $V^{\top}
V = I$. We thus conclude
$$X (X^{\top} X)^{-1} X^{\top} = V (V^{\top} V)^{-1} V^{\top} = V V^{\top}$$
as desired.
% Therefore, using the fact that \(U U^\top = I\) and \(V^\top V = I\),
% we have
% \[(V |D|^{1/2}) ((V |D|^{1/2})^\top (V |D|^{1/2}))^{-1} (V |D|^{1/2})^\top = 
% (X U Q) ((X U Q)^\top (X U Q))^{-1} (X U Q)^\top.\]
% The right-hand side becomes
% \[\begin{split}
%   (X U Q) ((X U Q)^\top (X U Q))^{-1} (X U Q)^\top & 
%     = X U Q Q^{-1} U^\top (X^\top X)^{-1} 
%     U (Q^\top)^{-1} Q^\top U^\top X^\top \\
%   & = X U U^\top (X^\top X)^{-1} U U^\top X^\top \\
%   & = X (X^\top X)^{-1} X^\top.
% \end{split}\]
% The left-hand side becomes:
% \[\begin{split}
%   (V |D|^{1/2}) ((V |D|^{1/2})^\top (V |D|^{1/2}))^{-1} (V |D|^{1/2})^\top & =
%     V |D|^{1/2} |D|^{-1/2} (V^\top V)^{-1} |D|^{-1/2} |D|^{1/2} V^\top \\
%   & = V V^\top.
% \end{split}\]

\textbf{Proof of Theorem \ref{theorem3}}. By Lemma \ref{lemma1},
\(V V^\top = X (X^\top X)^{-1} X^\top\) where \(X\) is defined as in
(\ref{eq:xy}). Since \(X\) is block diagonal with each block
corresponding to one community, \(X (X^\top X)^{-1} X^\top\) is also a
block diagonal matrix with each block corresponding to a community and
zeros elsewhere. Let $B = n VV^{\top}$. Then $B_{ij} = 0$ whenever
vertices $i$ and $j$ belong to different communities.  
%
%then the \(ij\)\textsuperscript{th} element of
%\(n X (X^\top X)^{-1} X^\top = n V V^\top = B\) is 0.
